A Computational Account of Default-Mode Function\\ by Control Theory and Reinforcement Learning

### 2016-09-26 paper/paper.tex
first intro draft

\section{The Default-Mode Network}
%
When left unperturbed, the human brain is not at rest. Rather, the brain continues to metabolize large quantities of oxygen and glucose energy to maintain inter-neuronal information transfer in the absence of an externally structured task (Bzdok and Eickhoff, 2016a). This baseline energy demand is subject to surprisingly little modulations due to the cognitive load posed by environmental challenges. What has early been described as the "stream of consciousness" in psychology (James, 1890).
%
A few years ago, the so-called "default mode network" was discovered entirely by accident (Shulman et al., 1997). A coherent set of brain regions consistently increased in neural activity during stimulus-independent thought, which was proposed to reflect the neural correlates underlying unfocused everyday mind wandering (Raichle et al., 2001). That is, in the beginning of the 21st century, brain imaging was the first technology to allow for the discovery of a coherent brain network that subserves a set of baseline mental activity (Bzdok et al., 2012; IBzdok et al., 2015).
%
15 years ago neuroscientists have discovered an energy-demanding brain network with unique properties. The "default-mode network" (DMN) appeared exclusive in task-induced deactivation during various psychological experiments. It was later argued to be systematically anti-correlated with brain regions that subserve task performance. Today, many authors speculate that the DMN implements continuous thinking about others’ and one’s own mind states, as well as adaptive envisioning of past, hypothetical, and future events. This network might have emerged to continuously predict environmental events using mental imagery as an evolutionary advantage.
%
Since 2001, the neurobiological properties of the default mode network have been investigated and reported in more than 3,000 neuroimaging publications 
The DMN was initially believed to represent the neural correlates of unconstrained mind-wandering because of its pervasively observed activity decrease during a large array of tasks. This processing of unknown information categories in the DMN has been argued to mediate an evolutionarily conserved function for the individual. This is all the more likely because the DMN contains the two biggest hotspots of energy consumption in the entire central nervous system. DMN activity also persists to a substantial degree during the early stages of sleep (Horovitz et al., 2008) and under anesthesia (Greicius et al., 2008). However, information processing in the DMN has also repeatedly been shown to impact human behavior. Goal-directed task performance improves with decreased activity in default-mode areas (Weissman et al., 2006) and increased DMN activity is linked to more task-independent, potentially disturbing thoughts (Mason et al., 2007). 
%
the DMN is likely to control the interplay between perception–action cycles and mental imagery (Bzdok et al., 2013a; Bzdok et al., 2013b). In particular, the DMN is likely to subserve a computational mechanism by its connections to highly associative brain areas that is key for human thought across time, splace, and content domains at the interface of external world and self (IBzdok et al., 2015).
%

hierarchically deep models of the world

Its societal and economic relevance is certified by involvement in various psychiatric disorders, including schizophrenia, autism, and depression.

sensory exchanges with the world

show how a formal approach can provide generic explanations for psycho- pathology that are physiologically grounded

quantitative (and parametric) characterisations offered by computational

d Biological systems are homeostatic (or allostatic), which
means that they minimize the dispersion (entropy) of their
interoceptive and exteroceptive states.
d Entropy is the average of surprise over time, which means
that biological systems minimize the surprise associated
with their sensory states at each point in time.
d In statistics, surprise is the negative logarithm of B

perception as hypothesis testing

Consider an organism that, in their present situation, is confronted by several choices of what to do next. Being able to accurately and richly mentally enact possible future states before making a decision would help to evaluate the desirability of different outcomes and also the planning processes needed to make them happen.

 the use of this constructive process goes far beyond simply predicting the future, to the general evaluation of fitness for purpose

scene construction processes when envisioning past, fictitious, and future events (Hassabis et al., 2007; Spreng et al., 2009).

It was observed that the more external stimuli are predictable, the more reflection processes become detached from the actual sensory environment and the more stimulus-independent, self-focused thoughts occur (Mason et al., 2007).

\begin{itemize}
  \item includes brain regions with highest baseline energy consumption in humans
  \item most active in non-disturbed mind-wandering without external influence
  \item tends to deactivate when humans engage in goal-direct tasks
  \item fluctuations in the DMN are related to lapses during and performance in
  such externally structured/focused tasks
  \item includes brain regions that come online first when waking up from
  anesthesia
  \item its network nodes are late to myelinate, an indicator of cognitive sophistication
  \item includes the most advanced processes hierarchies and has
  no direct connections with sensory regions processing external input
  \item the DMN's conceivable key role in the continuous environmental
  tracking in a generative, integrative process might explain both its highest energy consumption in the brain and its intimate coupling with conscious awareness
  \item Indeed, patients with right IPL (=DMN) damage have particular difficulties
  with multistep actions (Hartmann et al. 2005)
  and with imagined action (Sirigu et al. 1996).
  \item Lesion of the hippocampus, feeding memory and spatial information, to the
  DMN impairs future and hypothetical thinking (cf. Hassabis PNAS)

  \item
  autobiographical memory sup- plies numerous building blocks of social semantic knowl- edge (Bar 2007; Binder et al. 2009). These isolated conceptual scripts may be reassembled to enable the fore- casting of future events (Tulving 1983, 1985; Schacter et al. 2007). Similar brain mechanisms in remembering past episodes and envisioning future circumstances is supported by their engagement of identical brain areas, as evidenced by a quantitative meta-analysis (Spreng et al. 2009). Moreover, retrograde amnesic patients were repor- ted to be impaired not only in prospection but also in imagining novel experience (Hassabis et al. 2007). These findings suggest a single neural network for mentally constructing plausible semantic scenarios of detached sit- uations regardless of temporal orientation (Buckner and Carroll 2007; Hassabis and Maguire 2007). Indeed, con- struction of detached probabilistic scenes has been argued to influence ongoing decision making by estimating out- comes of behavioral choices (Boyer 2008; Suddendorf and Corballis 2007; Schilbach et al. 2008b). 
  \item
  DMN exists in monkeys [73] and rats [74]. 
  \item clinical research has corroborated default-mode network dysfunction in various psychiatric and neurological disorders
\end{itemize}

The DMN poses an unprecedented challenge to our neuroscientific understanding and methodological inventory.

Despite ramifications in health and disease, the functional role of the DMN poses an unprecedented challenge to the conventional interpretation strategies and existing methodological arsenal in the neurosciences.

Represents the physiological and pathological instantiation of a human beings' default mental repertoire, the nature of which remains largely obscure.

The DMN has now urged major attention in the neuroscientific community (Raichle et al., 2001, >6600 citations).

\section{Control Theory and Reinforcement Learning}
\paragraph{Control Theory.}
\begin{itemize}
\item mathematically formalized dynamics systems with input, output and
feedback loops that produce error signals for adaptation
\item typically realized by (non-linear) differential equations
\item has been used to describe neural systems in the past
\item gradient update = learn/adapt system dynamics by error occurrence
\item related to notions of game theory
\item switching between states -> typically reduces to a sequential processing model
\item Tenenbaum2011Science: Much recent work points to Monte Carlo or stochastic sampling– based approximations as a unifying framework for understanding how Bayesian inference may work practically across all these levels, in minds, brains, and machines (70–74).
\item policy: action to take given a state
\item value: prediction of future reward
\item model: predict what the environment will do next
\item model-baseed RI learning: we try to understand the environment; it is initially unknown
\item special to RL: exploitation vs. exploration
\item Tenenbaum2011Science: Much recent work points to Monte Carlo or stochastic sampling– based approximations as a unifying framework for understanding how Bayesian inference may work practically across all these levels, in minds, brains, and machines (70–74).
\item Tenenbaum2011Science: Much recent work points to Monte Carlo or stochastic sampling– based approximations as a unifying framework for understanding how Bayesian inference may work practically across all these levels, in minds, brains, and machines (70–74).
\item Tenenbaum2011Science: Much recent work points to Monte Carlo or stochastic sampling– based approximations as a unifying framework for understanding how Bayesian inference may work practically across all these levels, in minds, brains, and machines (70–74).
\end{itemize}

\section{Known Neurobiological Properties of the Default-Mode Network}
mPFC and PCC = highest metabolic consumption in the brain/highest level of basal glucose energy consumption in humans

scene con- struction processes when envisioning past, fictitious, and future events (Hassabis et al., 2007; Spreng et al., 2009; Bzdok et al., 2013).
it is not random but purposeful
draws on conceptual knowledge acquired through experience
operating on internal sources of information
Computational processes with a focus on the internal world,
when the environment is highly predictable and
external stimuli are irrelevant.
involvement of the PMC across the boundaries of classical cognitive domains

mechanistic relevance of each node -> we deconstruct

PFC
Patients with prefrontal lesions have problems in adapting to novel situations and stimuli
Stuss, D. T. \& Benson, D. F. The Frontal Lobes (Raven, New York, 1986).

vmPFC regions respond more to the value difference that will be acted on in the current trial (executed in red), and dmPFC to the currently irrelevant value difference (modeled in blue) {Nicolle, 2012 \#3750} individual. Instead activity in vmPFC reflects a choice preference that is executed and rostral dmPFC a choice preference that is modeled. Thus, dorsal regions of rostral mPFC contain value representations in the frame of reference of a modeled (as opposed to an actuated) preference irrespective of whether this applies to another’s or to one’s own likely actions and goals. Likewise, ventral mPFC contains a representation of value in the frame of reference of an executed choice, even if this executed choice reflects one’s own or another’s preferences. dmPFC that reflects the values and preferences of another individual (here temporally discounted at a rate specific to the individual), even when they are not directly relevant to the task at hand.

Simony 2016
This implies that the processing timescale of the DMN is longer than the length of single paragraphs (B40s) and that information accumulated from prior paragraphs can affect ongoing processing in this system
Our results suggest that the DMN accumulates and integrates information over minutes. Stimulus-induced correlation patterns in the DMN were largest for the intact story, greatly decreased in the paragraph scramble condition, and were almost entirely abolished in the word scramble condition.

@@

strategic selection (Pearson et al., 2009), risk assessment (McCoy and Platt, 2005), and outcome-contingent behavioral modulation (Hayden et al., 2008),

@@

@@

The dmPFC subserves predominantly ambiguous amodal processes
across time, space, and content domains in
sensory-independent top-down pathways.
This DMN node has been described as a “mental sketchpad” (Goldman-Rakic, 1996),
potentially implicated in de-novo generation and binding
of meaning representations instructed by stored semantics and memories.
%
The dmPFC may thus enable inference, representation, and assessment
of one's own and other individuals' action and thoughts.
For instance, neural activity in human dmPFC
reflected expectations about other peoples actions and errors thereof
(Suzuki et al., 2012; van Schie et al., 2004).
dmPFC activity indeed explained the proficiency decline
of inferring other peoples thoughts in the elderly (Moran et al., 2012).
Some dmPFC neurons in macaque monkeys exhibited a preference
for processing others', rather than own, behavior
with fine-grained adjustment of contextual circumstances (Yoshida et al., 2011).
%
Such highly abstract neural computations necessarily rely on the
generation of probabilistic internal information drawing from
episodic memory retrieval, non-mnemonic scene construction processes,
and explicit knowledge of the external world.

@@

Finally,
the right and left TPJ are known to have hemispheric differences
according to their cytoarchitectonic borders and gyrification pattern
(Caspers et al., 2008; Caspers et al., 2006).
Neuroscientific investigations on hemispheric functional specialization
have converged to the right versus left cerebral hemisphere as dominant for
attention versus language functions (Broca 1865, Wernicke 1874, Lichtheim 1885;
Gazzaniga, Bogen et al. 1965, Sperry 1982, Stephan, Marshall et al. 2003).
%
The aRTPJ is a key node network is central for
action initiation during externally structured tasks and
sensorimotor control by integrating supramodal stimulus-guided attention
(Corbetta et al., 2002).
Involvement of this DMN node was repeatedly reported in
visuo-proprioceptive conflict (Balslev et al., 2005) and
multi-modal detection of sensory changes across
visual, auditory, or tactile stimulation in a multi-modal fMRI study
(Downar et al., 2000).
In humans, direct electrical stimulation of the
RTPJ during neurosurgery was associated with altered perception
and stimulus awareness (Blanke et al., 2002).
%
Importantly, the RTPJ has been shown to be intimately related to
supramodal prediction error signaling.
Neural activity in the RTPJ has been argued to be responsible 
for stimulus-driven attentional reallocation to 
salient and surprising sources of information
as a ‘‘circuit breaker’’ that recalibrates control and maintenance systems
(Bzdok et al., 2013; Corbetta et al., 2008).
In the face of large discrepancies between actual and previously predicted
environmental events the RTPJ acts a potential switch between
externally-oriented mind sets focussed on the
sensory world and internally-oriented mind sets focussed
on self-relevant mind-wandering of possible mental scenarios.
Transient RTPJ in humans for instance disrupted the
impact of predicted intentions of other individuals (Young et al. 2010b),
a capacity believed to be subserved by the DMN.
The RTPJ might hence be an important player that shifts away
from the default active ‘‘internally directed’’ processes
to deal instead with immediate environmental objects and contexts.
-> (slow) endogeneous changes in state transitions by the policy and value matrices
For example, the alarm at the museum is not part of the task set of listening to the guide’s discussion of Hieronymous Bosch, but it is clearly a behaviourally relevant stimulus. <=> MCMC in HC / PCC

@@

>>>>

\section{Introduction}
%
When left unperturbed, the human brain is not at rest.
Rather, the brain continues to metabolize large quantities of
oxygen and glucose energy to maintain inter-neuronal information
transfer in the absence of concrete behavioral goal (Bzdok and Eickhoff, 2016a).
This baseline energy demand is subject to surprisingly little modulations
when mentally processing environmental challenges.
What has early been described as the "stream of consciousness"
in psychology (James, 1890), found its neurobiological manifestation 
in the so-called "default mode network" (DMN) was discovered entirely by accident
(Shulman et al., 1997). This regrouping of some of the highest regions in
the human association cortex (Mesulam et al., 1998) consistently
increased in neural activity during
unfocused everyday mind wandering (Raichle et al., 2001). That is, in
the beginning of the 21st century, brain imaging was the first technology
to allow for the discovery of a unique brain network that subserves a
set of baseline mental activity (Bzdok et al., 2012; Bzdok et al., 2015).

The "default-mode network" (DMN) appeared to be exclusive in task-induced
deactivation during various psychological experiments.
This processing of unknown information categories in the DMN has
been argued to mediate an evolutionarily conserved function for the individual.
This is all the more likely because the DMN contains the two
biggest hotspots of energy consumption in the entire central nervous system.
DMN activity also persists to a substantial degree during the early stages of
sleep (Horovitz et al., 2008) and under anesthesia (Greicius et al., 2008).
Today, many authors believe that the DMN implements
probabilistic estimation of past, hypothetical, and future events. This
brain network
might have emerged to continuously predict environmental events using
mental imagery as an evolutionary advantage.
%
However, information processing in the DMN has also repeatedly
been shown to directly impact human behavior. Goal-directed task performance
improved with decreased activity in default-mode areas (Weissman et al., 2006)
and increased DMN activity was linked to more task-independent,
potentially disturbing thoughts (Mason et al., 2007). 
%
The overarching function of the DMN is mystified by
controlling the interplay between perception–action cycles and
maintaining a baseline contemplations
across time, splace, and content domains at the interface of external world
and self (Bzdok et al., 2015).

Consider an organism faced by choosing the next action
and being informed by rich
reenactment of really happened, hypothetically conceived, and
predicted future events to perfect ongoing decision making.
Control theory is a particularly attractive computational framework
to describe, quantify, and predict the autonomously acting systems,
perhaps including the human brain.
It is known that, the more external world is predictable,
the more mental activity becomes detached from the actual sensory environment
and the more stimulus-independent thoughts occur
(Mason et al., 2007)
that may contribute to optimized control of the organism.
A policy matrix captures the possible action than can be performed
on the environment given a current state. A value matrix 
maps maps expected rewards to environmental objects and events.
Switching between states reduces to a sequential processing model.
In a model-free reinforcement learning approach,
we try to understand the environment always better
starting, at birth, with an environment that initially unknown.
Informed by gradient updates,
the adaptive DMN dynamics learned by feed-back loops
that are calibrated by error occurrence.
Reinforcement learning frameworks can naturally embed DMN function
into the tension between exploitative action with immediate reward and
explorative action with longer-term reward schedules.
The diversity of DMN implications in human cognitive processes
can be parsimoneously recast as minimizing prediction error
and maximizing reward outcome at various time scales.
Such a biologically meaningful optimization objective
may be solved by a stochastic approximation
based on a brain implementation of Markov Chain Monte Carlo sampling
(Tenenbaum2011Science). Even (necessarily imperfect) memories
of past experience, random mind-wandering and dreams during sleep
may be purposeful as building blocks to iteratively improve
the predictive DMN machinery to optimize behavior of the organism.

\section{Known Neurobiological Properties of the Default-Mode Network}
We will attempt to deconstruct the overall DMN function
into the mechanistic relevance of each individual network node
by reviewing experimental neuroscience evidence.
The PMC and the dmPFC within the DMN
are responsible for the highest metabolic turn-over
of glucose energy consumption in humans (Raichle).

@@

strategic selection (Pearson et al., 2009), risk assessment (McCoy and Platt, 2005),
and outcome-contingent behavioral modulation (Hayden et al., 2008),

@@

Also in humans, 
the DMN has been shown to gather and integrate information
over different paragraphs in an fMRI study on auditory narratives (Simony 2016).

@@

The dmPFC subserves predominantly ambiguous amodal processes
across time, space, and content domains in
sensory-independent top-down pathways.
This DMN node has been described as a “mental sketchpad” (Goldman-Rakic, 1996),
potentially implicated in de-novo generation and binding
of meaning representations instructed by stored semantics and memories.
%
The dmPFC may thus enable inference, representation, and assessment
of one's own and other individuals' action and thoughts.
Patients with neurological lesions in the prefrontal cortex
are now to adapt to novel situations and stimuli with difficulty
(Stuss, D. T. \& Benson, D. F. The Frontal Lobes, Raven, New York, 1986).
Specifically, neural activity in human dmPFC
reflected expectations about other peoples actions and errors thereof
(Suzuki et al., 2012; van Schie et al., 2004).
dmPFC activity indeed explained the proficiency decline
of inferring other peoples thoughts in the elderly (Moran et al., 2012).
Some dmPFC neurons in macaque monkeys exhibited a preference
for processing others', rather than own, behavior
with fine-grained adjustment of contextual circumstances (Yoshida et al., 2011).
%
Such highly abstract neural computations necessarily rely on the
generation of probabilistic internal information drawing from
episodic memory retrieval, non-mnemonic scene construction processes,
and explicit knowledge of the external world.
%
According to computational neuroimaging experiments,
the dmPFC activity preferentially models stimulus-value associations of
possible actions that are not actually executed,
whereas the vmPFC activity models value in the environment that leads performed actions
{Nicolle, 2012 \#3750}.

@@

Finally,
the right and left TPJ are known to have hemispheric differences
according to their cytoarchitectonic borders and gyrification pattern
(Caspers et al., 2008; Caspers et al., 2006).
Neuroscientific investigations on hemispheric functional specialization
have converged to the right versus left cerebral hemisphere as dominant for
attention versus language functions (Broca 1865, Wernicke 1874, Lichtheim 1885;
Gazzaniga, Bogen et al. 1965, Sperry 1982, Stephan, Marshall et al. 2003).
%
The aRTPJ is a key node network is central for
action initiation during externally structured tasks and
sensorimotor control by integrating supramodal stimulus-guided attention
(Corbetta et al., 2002).
Involvement of this DMN node was repeatedly reported in
visuo-proprioceptive conflict (Balslev et al., 2005) and
multi-modal detection of sensory changes across
visual, auditory, or tactile stimulation in a multi-modal fMRI study
(Downar et al., 2000).
In humans, direct electrical stimulation of the
RTPJ during neurosurgery was associated with altered perception
and stimulus awareness (Blanke et al., 2002).
%
Importantly, the RTPJ has been shown to be intimately related to
supramodal prediction error signaling.
Neural activity in the RTPJ has been argued to be responsible 
for stimulus-driven attentional reallocation to 
salient and surprising sources of information
as a ‘‘circuit breaker’’ that recalibrates control and maintenance systems
(Bzdok et al., 2013; Corbetta et al., 2008).
Indeed, patients with right TPJ damage have particular difficulties
with multistep actions (Hartmann et al. 2005).
In the face of large discrepancies between actual and previously predicted
environmental events the RTPJ acts a potential switch between
externally-oriented mind sets focussed on the
sensory world and internally-oriented mind sets focussed
on self-relevant mind-wandering of possible mental scenarios.
Transient RTPJ in humans for instance disrupted the
impact of predicted intentions of other individuals (Young et al. 2010b),
a capacity believed to be subserved by the DMN.
The RTPJ might hence be an important player that shifts away
from the default active ‘‘internally directed’’ processes
to deal instead with immediate environmental objects and contexts.
-> (slow) endogeneous changes in state transitions by the policy and value matrices
For example, the alarm at the museum is not part of the task set of listening to the guide’s discussion of Hieronymous Bosch, but it is clearly a behaviourally relevant stimulus. <=> MCMC in HC / PCC

@@

At the least, we propose an alternative vocabulary to
describe and discuss neuroscientific studies on the DMN.
